---
title: "Writing a great story for data science projects - Spring 2026 "
subtitle: "This is a Report Template Quarto"
author: "Student Name (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

Nice report!
:::

## Introduction

Insert interesting hook from Elizabeth's research... In a world filled with intense socio-economic and geo-political strife, it is no wonder that the people of the planet are facing the mental health crisis of the millennium. The temptation to escape or even just cope is one many cannot overcome. This is the "Information age"-- perhaps a more appropriate title: the over-information age. Whether the pressure arises from forces external: employment, bills, family, or from within, Dr. Reddit and TikTok therapists are quick to offer up the next "hack" for your body or mind. Drugs to solve any problem abound, a person is spoiled for choice... but what choice will a person make?

Can we capture this information and utilize it in constructively? What if we used this to ease the human condition instead of criminalize it?

The dataset we have chosen to model is: 

We deigned to design a model that can be used not only to predict what types of drugs a person might be inclined to use but also reflexively- this time using a person's "drug of choice" to inform them of potential underlying psychological conditions or challenges they may be able to seek proper pharmaceutical help to mitigate. The War on Drugs failed, its time to work smarter, not harder.  


<!-- -->

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

What better tool to predict the decisions from a human brain than by an algorithm designed to work like one: the neural network. In our case, we explored both traditional Deep Neural Networks (DNNs) like multilayer perceptrons as well as Recurrent Neural Networks (RNNs). 

The goal of a neural network has always been straightforward: make the computer work like a brain: a complex, multi-parallel system of individual processing units, "neurons". [@MontesinosLopez2022] The inception of the artificial neuron goes back as far as the early 1940's and the underlying mechanism has remained fundamentally unchanged since. Artificial Neural Networks (ANNs) of the era were "simple" consisting of input layer, activation layer, and output layer. The largest hurdle to be overcome being a lack of processing and optimizing ability. Due to these structure and computational limits, these models were only suited for "simpler" linear problems. *though honestly, constructing and executing a neural network in the 40s sounds to us as anything BUT simple* [@ekundayo2025deep] Neural networks enter  a period of stagnation until the year 1986 with the introduction of the "novel method" of back propagation.[@Rumelhart1986] Flash forward 40 years and the ANN has all but inundated all corners of data analysis. 

The appropriate "version" of ANN to select is largely based on the input data at hand. Are your data points depended on one another or dependent on time itself? An RNN is best suited to this processing. Do you intend to decipher and interpret imagery? Best look into CNNs. But why? And furthermore, what version should we use for this drug dataset? 

EXPLAIN THE STRUCTURE OF A TRADITIONAL DNN

EXPLAIN THE STRUCTURE OF A TRADITIONAL RNN

EXPLAIN THE STRUCTURE OF A TRADITIONAL CNN

Traditional feed forward DNNs process data in a one way flow. A human brain, in contrast is a highly interconnected network with feedback loops and memory, much like the structure of an RNN. Our model is attempting to replicate the function of the brain however, the data is completely comprised of continuous or categorical non-sequential observations. To rectify this incongruity and leave nary a stone unturned, we explored both a traditional DNN and an RNN.


## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
