---
title: "Predicting Drug Use Using Personailty Type and Pyschological Diagnoses- Summer 2026 "
subtitle: "A comparasion between RNN and DNN on non-sequential data"
author: "Elizabeth Rose and Victoria Kipker (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
engine: knitr
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

<style>
body {
  text-align: justify;
}
</style>

## Introduction

  In an era plagued with a severe mental health crisis accelerated by both socioeconomic strife and the cognitive overload of the information age, the temptation to cope or escape can be compelling. As individuals attempt to navigate the current global landscape, many seek assistance from unregulated sources online, such as social media, frequently leading to self-diagnosis and self-medication. This access to “over-information” and a wide variety of ever-evolving psychoactive substances available have essentially left individuals overwhelmed by choice. Consequently, a critical behavioral question remains: what innate factors influence an individual’s substance of choice?

  To investigate this phenomenon, research conducted by Fehrman et al. explored individual’s risk for drug consumption using a dataset collected via an online survey. [@fehrman2017five] Administered over a 12-month period, the survey resulted in usable data from 1,885 participants across a range of English-speaking nations including the United Kingdom, the United States, Canada, Australia, New Zealand, and Ireland. In addition to collecting information regarding the use of 18 legal and illegal drugs, the survey collected demographic metrics, such as education level and race, as well as their personality traits.


  Personality traits were assessed using the Revised NEO Five-Factor Inventory (NEO-FFI-R). This method evaluates personalities based on scores across five core dimensions: neuroticism, extraversion, openness to experience, agreeableness, and conscientiousness [@costa1992revised]. This dataset established by Fehrman et al. serves as the foundation for this study. This study attempts to model the complex relationship between personality profiles and an individual’s “drug of choice”, while simultaneously predicting substance risks by implementing both traditional Deep Neural Networks (DNNs), such as the multilayer perceptron, as well as Recurrent Neural Networks (RNNs). 


<!-- -->

  The concept for the function of artificial neural networks (ANNs) has always been to replicate the processing methods of the brain: a complex, multi-parallel system of individual processing units called “neurons” [@MontesinosLopez2022] [@Larochelle2009Exploring]. While the inception of the artificial neuron dates back to the early 1940s, the underlying mechanism has remained fundamentally unchanged. Early iterations of ANNs were structurally simple, consisting of a single input layer, activation function, and an output layer. At the time, these models were operating with both structural and computational limitations, and as such were restricted to solving simpler, linearly separable problems [@ekundayo2025deep]. The field then experienced a prolonged period of stagnation until it ended abruptly in 1986 with the introduction of backpropagation. This optimization method allows networks to systematically and repeatedly adjust their weights across multiple hidden layers to minimize the difference between the actual output and the desired output [@Rumelhart1986]. In the decades since the popularization of backpropagation, advancements in computing power and data availability have accelerated the evolution of deep learning, to the point where ANNs have permeated virtually all corners of modern data analysis. 

## Methods

  The multilayer perceptron (MLP) remains the standard feedforward architecture for processing structured, tabular data. Within an MLP, information can only flow in a single direction from the input layer through to the hidden processing layer or layers. Notably, neurons within the same layer are not interconnected. [@elmandouh2026comparing]. While more advanced models have been developed, MLPs have remained the benchmark when both simplicity and computational efficiency are the primary objectives. Furthermore, many optimization techniques have been developed to fine-tune these models, significantly accelerating convergence rates during training [@przybyla2024multi].
	
  The selection of an appropriate ANN is dictated by the structure of the input data. For time series data, a recurrent neural network (RNN) provides increased analytical power by incorporating short-term memory. Unlike the unidirectional flow within an MLP, an RNN allows information to propagate through feedback loops. This enables the neurons to have connections to the previous layer, connections to neurons in the subsequent layer, and connections that transmit information between other neurons within the same layer. Essentially, this built-in memory enables the model to maintain context across a sequence to link current inputs to historical information [@elmandouh2026comparing].

  For the analysis of time series data, language data, or data with a spatial structure, such as images or videos, a convolutional neural network (CNN) is typically the most effective. These models function by applying a series of convolutional filters that perform hierarchical feature extraction. Initially, low-level details such as edges or points are analyzed, while subsequent layers analyze higher-level shapes and contours. Since CNNs utilize the spatial correlation between adjacent data points, data generally requires less preprocessing than more traditional methods [@elmandouh2026comparing]. 

  For the purpose of this paper, our analysis will focus exclusively on MLP and RNN models. By comparing the feed-forward processing of the MLP against the feedback-oriented structure of the RNN, we aim to determine if personality traits are good predictors of drug use. 



The models we have created utilize two layers of relu activation funtions followed by a sigmoid activated output layer.
the input and hidden layers utilize the ReLU or rectified linear unit function, . 


  The MLP model developed for this analysis consists of an input layer, two hidden layers, and an output layer. The first hidden layer has 64 neurons and uses Rectified Linear Unit (ReLU) activation, where postiive values are passed through as identity and negative values are rendered as 0, while the second hidden layer has 32 neurons and also uses ReLU activation. The use of this activation function is computationally inexpensive and assists with mitigating vanishing gradient issues that arise with the use of tanh or sigmoid functions. Additionally, even thought the ReLU function will inveitably render some neurons inactive with a 0 output, we have also employed a dropout parameter which ensures netowrk sparsity, making sure that neurons in the net do not become co-dependent. The output layer consists of 15 neurons and uses a sigmoid activation since we will be conducting a multi-label classification task.

  Mathematically, these layers can be explained as seen below:

Hidden layer 1 Calculation with M1 neurons 
$$ V1_j = g_1\left(\sum_{i=1}^{d} w^{(1)}_{ji} x_i\right) for\ j = 1, ...,M_1 $$

Hidden layer 2 Calculation with M2 neurons 
$$ V2_k = g_2\left(\sum_{j=1}^{M1} w^{(2)}_{kj} V1_j\right)for\ k = 1, ...,M_2 $$
Output layer calculation 
$$ y_l = g_3\left(\sum_{k=1}^{M2} w^{(3)}_{lk} V2_k\right)for\ l = 1, ...,M_3 $$
  Here, the first hidden layer processes the input features x using weights w(1). The second hidden layer then processes the output of the first hidden layer V1. Finally, the output layer is used to compute the predicted drug usage probabilities y [@MontesinosLopez2022]. The activation functions at each layer are represented by g1, g2, and g3. As previously mentioned, ReLU will be used for the hidden layers in our MLP model. This activation function performs a nonlinear transformation by acting as a switch. If the input is less than 0, then the output will be 0. However, if the input is above a certain threshold, the output is the identify of the input.

ReLU activation function: 
$$f(z) = \begin{cases} 0 & \text{for } z < 0 \\ z & \text{for } z \ge 0 \end{cases}$$
 The choice of output layer activation function was between sigmoid, where the probabilities of output classes are independen or the softmax equation which produces probabilities of classes that sum to 1.  It is our belief that the sigmoid activation function best represents the human behavior of being equally (highly) likely to engage in multiple drugs, that a person could have multiple independent high proabbilites of substance use (e.g. a person could have a 90% probability of using cocaine and an 85% probability of using meth).

Sigmoid activation function
$$f(z) = \frac{1}{1 + e^{-z}}$$  
Softmax activation function
$$P(y = i \mid \mathbf{z}) = \frac{e^{z_i}}{\sum_{j=1}^{K} e^{z_j}}$$
The RNN model for this analysis consists of one hidden layer and a dense, or output, layer. The hidden layer is a limited short-term memory (LSTM) layer of 64 neurons that houses intermediates. The output layer then consists of 15 neurons and a sigmoid activation function, one per drug we are predicting. 

The LSTM employs the use of gates to process information. There are three gates, a forget gate, which decides what previous information is no longer relevant to the processing, the input gate which filters incoming information and determines what is valuable to be saved and passed along, and the output gate which transforms information gathered and processed/filtered into a decision. These gates are mathematically described below: 

forget gate: 
$$f_t = \sigma(W_f x_t + U_f h_{t-1} + b_f)$$

input gate: 
$$i_t = \sigma(W_i x_t + U_i h_{t-1} + b_i)$$
output gate: 
$$o_t = \sigma(W_o x_t + U_o h_{t-1} + b_o)$$

```{r }

```


### Data Exploration and Visualization

  The dataset used in this analysis was collected via an online survey by Fehrman et al. [@fehrman2017five] initially yielded results from 2,051 individuals. To ensure data integrity, a fictitious drug “Semeron” was included as a validity check to identify inattentive respondents and individuals inflating their drug use. Information from individuals who failed this validity check or incorrectly completed the survey was removed, resulting in a usable sample size of 1885 participants. The majority of the respondents resided in the United Kingdom (55.4%) and the United States (29.5%), with the remainder from Canada, Australia, Ireland, and New Zealand. Additional demographic information included age, gender, education level, and ethnicity.

  To assess personality traits, participants completed the Revised NEO Five-Factor Inventory (NEO-FFI-R), responding to 60 statements on a 5-point Likert scale. Impulsiveness was measured using the Barratt Impulsiveness Scale$$f_t = \sigma(W_f x_t + U_f h_{t-1} + b_f)$$

$$i_t = \sigma(W_i x_t + U_i h_{t-1} + b_i)$$

$$o_t = \sigma(W_o x_t + U_o h_{t-1} + b_o)$$ (BIS-11), which assesses three sub-factors: motor, attentional, and non-planning impulsiveness. Additionally, sensation-seeking tendencies were evaluated using the Impulsive Sensation Seeking (ImpSS) scale, consisting of  19 true/false statements. 

  Regarding substance use, participants reported their use of 18 legal and illegal drugs. For each substance, respondents selected from seven usage frequency categories: never used, used over a decade ago, used in the last decade, year, month, week, or day. These categorical responses were then quantified into numerical values by the original researchers, allowing the data to be processed by both statistical and machine learning models as continuous features. 

  We begin exploration of this dataset by examining the distribution of reported drug use. See Figure 1 below

```{r}
library(readr)

# Reading the file
library(ucimlrepo)
drug_data = fetch_ucirepo(id=373)
data = drug_data$data$original
#head(data)  

# Demographic reatures have already been scaled (Z-score standardization), however drug columns need to be converted
# Rows where individuals claimed "semer" use need to be removed since it was included as a fake drug


library(dplyr)

# Vector of all the drug column names
drug_cols = c(
  "alcohol", "amphet", "amyl", "benzos", "caff", "cannabis", "choc", 
  "coke", "crack", "ecstasy", "heroin", "ketamine", "legalh", "lsd", 
  "meth", "mushrooms", "nicotine", "semer", "vsa"
)

target_drugs = c(
  "amphet", "amyl", "benzos", "cannabis",
  "coke", "crack", "ecstasy", "heroin", "ketamine", "legalh", "lsd", 
  "meth", "mushrooms", "nicotine", "vsa"
)

# Now we filter out "Semeron" participants by keeping ones that say "never used"
data_cleaned = data %>%
  filter(semer == "CL0")


# Converting drug consumption columns into a binary target (user vs non-user)
# 0(non-user) and 1 (user)
## We will say anyone who has used the particular drug within the last year is a "user"

data_all_binary = data_cleaned %>%
  mutate(across(
    .cols = all_of(drug_cols),
    .fns = ~ ifelse(. %in% c("CL3", "CL4", "CL5", "CL6"), 1, 0)
  ))


# Checking to make sure they are binary
#head(data_all_binary[, c("alcohol", "amphet", "amyl")])


# Now we need to convert the features to a numeric matrix (X)
# Since we only want to train on personality scores, not other drug use, we need to remove other drug columns, too. 
X_data = data_all_binary %>%
  select(nscore, escore, oscore, ascore, cscore, impuslive, ss) %>%
  as.matrix()
#dim(X_data)


# Extracting the target variables (drug columns)
y_matrix = data_all_binary %>%
  select(all_of(target_drugs)) %>%
  as.matrix()
# Verifying the shape
#print(paste("Features Matrix (X) shape:", paste(dim(X_data), collapse = "x")))
#print(paste("Multi-Label Target Matrix (y) shape:", paste(dim(y_matrix), collapse = "x")))
y_matrix1 = data_all_binary %>%
  select(all_of(drug_cols)) %>%
  as.matrix()

# Plotting class imbalance 
library(ggplot2)
library(tidyr)
library(dplyr)

# Converting matrix to a dataframe
imbalance_plot_data = as.data.frame(y_matrix) %>%
  gather (key = "Drug", value = "Status") %>%
  mutate (Status = ifelse(Status == 1, "User", "Non-User"))

imbalance_plot_data1 = as.data.frame(y_matrix1) %>%
  gather (key = "Drug", value = "Status") %>%
  mutate (Status = ifelse(Status == 1, "User", "Non-User"))

# Creating a bar chart
ggplot(imbalance_plot_data1, aes (x = reorder(Drug, Status == "User", FUN = sum), fill = Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Non-User"= "gray", "User" = "blue"))+
  coord_flip() +
  labs(
    tag = "Figure 1",
    title = "Class Imbalance Across All 18 Target Substances",
    x = "Substance",
    y = "Percentage of Dataset",
    fill = "Classification"
  ) +
  theme_minimal()

```
  Based on Figure 1, we noticed two facts about the data: there is a large class imbalance in reported drugs used (very few respondents reporting usage of drugs like crack) and this may cause the model to struggle in predicting use of those, and there are three substances that most if not all respondents reported use of: choc (chocolate), caff (caffeine), and alcohol. This becomes important when we attempt to rectify class imbalances before splitting into train/test groups. 

  A random split without taking into account this class imbalance can compound the imbalance impact. In order to mitigate this, we introduce the use of a composite score. A sum of substances used is calculated for each of the instances in the dataset then taken into consideration for the train/test split purposes (this results in respondents reporting multiple substances used are balanced between train and test). 

  We decided to leave out the three most common substances used to make the composite score more reflective of illicit or multi-drug use. This way a person reporting only the use of those three would now have a composite score of 0 instead of 3 which may be the same as a person who reported use of three illicit drugs. 

```{r}
# Since we have imbalance for the majority of the drugs, we need to be careful with our train/test split. 
# A random split will just compound the imbalance issue.
# We will use a composite profile score 
library(caret)
set.seed(10)

# By using this method, individuals that use more than one drug will be evenly distributed
composite_score = as.factor(rowSums(y_matrix))

# Conducting the 80/20 split
train_indices = createDataPartition(composite_score, p = 0.80, list = FALSE)

X_train = X_data[train_indices,]
X_test = X_data[-train_indices,]
y_train = y_matrix[train_indices,]
y_test = y_matrix[-train_indices,]

# Testing out the dimensions 
#print(paste("X_train:", paste(dim(X_train), collapse = " x ")))
#print(paste("y_train:", paste(dim(y_train), collapse = " x ")))
#print(paste("X_test:", paste(dim(X_test), collapse = " x ")))
#print(paste("y_test:", paste(dim(y_test), collapse = " x ")))

ggplot(imbalance_plot_data, aes (x = reorder(Drug, Status == "User", FUN = sum), fill = Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Non-User"= "gray", "User" = "darkgreen"))+
  coord_flip() +
  labs(
    tag = "Figure 2",
    title = "Class Imbalance Across 15 Target Substances",
    x = "Substance",
    y = "Percentage of Dataset",
    fill = "Classification",
  ) +
  theme_minimal()
```


  Figure 3 illustrates the split of users between the train and test groups. A fairly even split of users was achieved with this composite score method. 



```{r}
# 1. Calculate the user percentages for the training set
train_dist <- as.data.frame(y_train) %>%
  gather(key = "Drug", value = "Status") %>%
  group_by(Drug) %>%
  summarize(User_Percentage = mean(Status == 1) * 100) %>%
  mutate(Split = "Train")

# 2. Calculate the user percentages for the test set
test_dist <- as.data.frame(y_test) %>%
  gather(key = "Drug", value = "Status") %>%
  group_by(Drug) %>%
  summarize(User_Percentage = mean(Status == 1) * 100) %>%
  mutate(Split = "Test")

# 3. Combine both distributions into a single dataframe
split_comparison <- rbind(train_dist, test_dist)

# 4. Create the side-by-side bar chart
ggplot(split_comparison, aes(x = reorder(Drug, User_Percentage), y = User_Percentage, fill = Split)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("Train" = "#2c3e50", "Test" = "#e74c3c")) +
  coord_flip() +
  labs(
    tag = "Figure 3",
    title = "Distribution of Drug Users Across Train and Test Splits",
    subtitle = "Verifying stratification balance (Omitting Alcohol, Caffeine, and Chocolate)",
    x = "Substance",
    y = "Percentage of Users (%)",
    fill = "Dataset Split"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 10)
  )
```
  We wanted to examine the typical personality types that were associated with some of these drugs to identify if there were strong overlaps that might impact the predictive ability of our models. Figures 4 and 5 illustrate a "psychological footprint" of a user of two of our target substances. The blue ring identifies a typical "non-user" and the red ring identifies a typical user. The gaps between the red ring and the blue ring at the individual personality points indicate an larger or smaller propensity for that particular trait. 

```{r}
library(dplyr)
library(tidyr)
# 1. Install remotes if you don't already have it
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# 2. Download and install ggradar from GitHub
remotes::install_github("ricardo-bion/ggradar")
library(ggradar) 

min_max_scale <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

# 2. Scale the raw respondent traits globally FIRST
data_scaled_global <- data_all_binary %>%
  mutate(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), min_max_scale))

# 1. Prepare aggregate averages for cannabis
footprint_cannabis <- data_scaled_global %>%
  group_by(cannabis) %>%
  summarise(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), mean)) %>%
  rename(group = cannabis) %>%
  mutate(group = ifelse(group == 1, "Cannabis User", "Non-User"))

colnames(footprint_cannabis) <- c("group", "Neuroticism", "Extraversion", "Openness", "Agreeableness", "Conscientiousness", "Impulsivity", "Sensation_Seeking")

# 3. Generate the Radar Chart explicitly using the new meth object
ggradar(
  plot.data = footprint_cannabis,
  grid.min = 0, grid.mid = 0.5, grid.max = 0.7,
  values.radar = c("0%", "50%", "70%"),
  group.point.size = 3, group.line.width = 1, legend.position = "bottom",
  
  # 1. SHRINK THE INTERNAL RADAR TEXT
  axis.label.size = 3.5,     # Shrinks trait names (Neuroticism, Impulsivity, etc.)
  grid.label.size = 3,       # Shrinks the percentage labels (0%, 50%, 100%)
  legend.text.size = 9       # Shrinks the legend text at the bottom
) + 
  labs(
    title = "The Psychological Footprint of Cannabis Consumption", 
    tag = "Figure 4"
  ) +
  # 2. SHRINK THE TITLES AND TAGS
  theme(
    plot.tag = element_text(size = 10, face = "bold"),           # Neat, smaller "Figure 5"
    plot.title = element_text(size = 12, face = "bold"),         # Smaller, crisp title
    legend.title = element_text(size = 9)                        # Smaller legend title
  )


# 2. Scale the raw respondent traits globally FIRST
data_scaled_global <- data_all_binary %>%
  mutate(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), min_max_scale))

# 1. Prepare aggregate averages for crack
footprint_crack <- data_scaled_global %>%
  group_by(crack) %>%
  summarise(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), mean)) %>%
  rename(group = crack) %>%
  mutate(group = ifelse(group == 1, "Crack User", "Non-User"))

colnames(footprint_crack) <- c("group", "Neuroticism", "Extraversion", "Openness", "Agreeableness", "Conscientiousness", "Impulsivity", "Sensation_Seeking")

# 3. Generate the Radar Chart explicitly using the new crack object
ggradar(
  plot.data = footprint_crack,
  grid.min = 0, grid.mid = 0.5, grid.max = 0.7,
  values.radar = c("0%", "50%", "70%"),
  group.point.size = 3, group.line.width = 1, legend.position = "bottom",
  
  # 1. SHRINK THE INTERNAL RADAR TEXT
  axis.label.size = 3.5,     # Shrinks trait names (Neuroticism, Impulsivity, etc.)
  grid.label.size = 3,       # Shrinks the percentage labels (0%, 50%, 100%)
  legend.text.size = 9       # Shrinks the legend text at the bottom
) + 
  labs(
    title = "The Psychological Footprint of Crack Consumption", 
    tag = "Figure 5"
  ) +
  # 2. SHRINK THE TITLES AND TAGS
  theme(
    plot.tag = element_text(size = 10, face = "bold"),           # Neat, smaller "Figure 5"
    plot.title = element_text(size = 12, face = "bold"),         # Smaller, crisp title
    legend.title = element_text(size = 9)                        # Smaller legend title
  )


```

  Examining the resulting polygons shows us that a user of cannabis is more likely to exhibit traits of sensation seeking and openness more than that reported in non-users. Additionally, the polygon for a crack user shows a larger disparity between non-users and users for the sensation seeking and impulsivity categories with a lower-than-non-user response for conscientiousness and agreeableness. Generally though, the red polygons trend in similar directions for users vs non-users. This raises the question of how correlated usage of drugs is in this dataset. 

  We use a correlation matrix (Figure 6) to illustrate this tendency towards drugs used together. Figure 6 confirms the strong ties between use of multiple substances. As the lines increase in darkness, the strength of the usage correlation increases. These correlations as well as the overlapping psychological footprint maps indicate that many respondents likely reported the use of multiple drugs. This information becomes important later when deciding on an activation function for our output level. Do we want to predict across all drugs where the probabilities add to 1? Our data indicate that use of these drugs is not mutually exclusive, thus we want to predict from 0 to 1 for each individual drug as a respondent can have a high probability of using two ore more drugs. 


```{r}
library(reshape2)
library(igraph)
library(ggnetwork)
library(ggplot2)

# 1. Calculate correlation matrix between all 15 remaining drugs
drug_corr <- cor(y_matrix) 
diag(drug_corr) <- 0 

# 2. Filter out weak connections
drug_corr[abs(drug_corr) < 0.25] <- 0 

# 3. Build the network object
network_structure <- graph_from_adjacency_matrix(drug_corr, weighted = TRUE, mode = "undirected")

# 4. SPREAD OUT THE NETWORK: Generate spaced coordinates using a scaling trick
set.seed(123) # For reproducibility

# We pass an inverse weight array (1 - correlation) or a fixed small weight 
# to force the Fruchterman-Reingold layout to push nodes further apart
spaced_layout <- layout_with_fr(
  network_structure, 
  niter = 1500, 
  weights = rep(0.1, ecount(network_structure)) # Lower weight = weaker springs = more spread out!
)

# 5. Convert to ggnetwork using our custom spaced coordinates
network_data <- ggnetwork(network_structure, layout = spaced_layout)

# 6. Plot the spacious layout
ggplot(network_data, aes(x = x, y = y, xend = xend, yend = yend)) +
  
  # CHANGES HERE: Map link color to the correlation weight and make lines thicker
  geom_edges(aes(color = weight), size = 1.2, alpha = 0.8) +
  
  # Gradient palette for the links (e.g., yellow to red for low to high correlation)
  scale_color_viridis_c(option = "plasma", name = "Correlation\nStrength", direction = -1) +
  
  # Uniform node coloring (a clean slate grey or dark blue so the links pop)
  geom_nodes(color = "gray", size = 6) +
  
  # Labels positioned neatly over the nodes
  geom_nodetext(aes(label = name), fontface = "bold", color = "black", size = 4) +
  
  theme_blank() +
  labs(
    title = "The Web of Substance Co-occurrence",
    subtitle = "Link colors reflect correlation strength (Threshold > 0.25)"
  ) +
  theme(
    legend.position = "right"
  )


```

## Training a model to predict drug use on personality features

  The below plots of accuracy and loss over epoch show that there is not a tremendous accuracy for our model. This is expected due to the number of different drug classes we are trying to generate probabilities for in one model. We can instead examine the F1 scores of each drug individually to see how well the model is classifying users vs non-users. Values close to 1 mean the model is predicting drug usage well. It is interesting to note that the model struggles the most with predicting nicotine users. 



```{r}
######## READY FOR MLP USE WITH THIS TRAIN/TEST DATA########

library(keras3)
library(tensorflow)
# install.packages("keras3")


# Input layer
inputs = layer_input(shape = c(7))

outputs = inputs %>%
  layer_dense(units = 64, activation = 'relu') %>%
  layer_dropout(rate = 0.3) %>%
  layer_dense(units = 32, activation = 'relu') %>%
  layer_dropout(rate = 0.2) %>%
  layer_dense(units = 15, activation = 'sigmoid')

mlp_model = keras_model(input = inputs, output = outputs)

# Compiling the model
mlp_model %>% compile(
  optimizer = 'adam',
  loss = 'binary_crossentropy',
  metrics = c('accuracy')
)

# Training the model
history = mlp_model %>% fit(
  x = X_train,
  y = y_train,
  epochs = 20, 
  batch_size = 32,
  validation_split = 0.2,
  verbose = 0
)
  
plot(history)

# Evaluating the effectiveness of this model using F1 score
library(yardstick)
# Converting props to binary
pred_probs = predict(mlp_model, X_test, verbose = 0)
pred_binary = ifelse(pred_probs > 0.5, 1, 0)
results = data.frame(truth = as.matrix(y_test, estimate = pred_binary))

# Calculating for each drug individually
eval_results = data.frame()

for( i in 1:ncol(y_test)){
  drug_name = target_drugs[i]
  temp_df = data.frame(
    truth = factor(y_test[, i], levels = c(0, 1)),
    estimate = factor(pred_binary[, i], levels = c(0, 1))
  )
  # Getting F1 score
  f1 = f_meas(temp_df, truth, estimate)
  eval_results = rbind(eval_results, data.frame(Drug = drug_name, F1 = f1$.estimate))
}
print(eval_results)
```


  Now that we have generated a model for predicting drug use, let's examine a heat map of participants and their predicted drugs used (Figure 7) and a corresponding plot of hit vs miss for the model when using the ground truth for these participants (Figure 8). 


```{r}

# Heat map for the first participant's likelihood profile
all_probs = predict(mlp_model, X_test, verbose = 0)

colnames(all_probs) = target_drugs

# Creating a data frame
df_probs = as.data.frame(all_probs)
df_probs$Participant = 1:nrow(df_probs)

df_long = df_probs %>%
  pivot_longer(cols = -Participant, names_to = "Drug", values_to = "Probability")

# Selecting the first 50 participants 
df_subset = df_long %>% filter(Participant <=50)

# Creating the heatmap
ggplot(data = df_subset,aes(x = Drug, y = factor(Participant), fill = Probability)) +
  geom_tile()+
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Individual Drug Usage Likelihoods",
       x = "Substance",
       y = "Participant ID", 
       tag = "Figure 7")


# 1. Convert probabilities to binary predictions (using 0.5 threshold)
all_preds = ifelse(all_probs >= 0.5, 1, 0)
colnames(all_preds) = target_drugs

# 2. FIX: Ensure y_test matches all_preds in dimension and order
# Convert y_test to a clean matrix and align columns with your target drugs
y_test_matrix = as.matrix(y_test[, target_drugs])

# Double check rows: If y_test is still the full dataset, slice it to match the test size
if (nrow(y_test_matrix) != nrow(all_preds)) {
  # This grabs the trailing rows matching your test set size
  y_test_matrix = tail(y_test_matrix, nrow(all_preds)) 
}

# 3. Calculate correctness element-by-element safely
is_correct = (all_preds == y_test_matrix)

# 4. Convert to data frame and add Participant IDs
df_correct = as.data.frame(is_correct)
df_correct$Participant = 1:nrow(df_correct)

# 5. Reshape to long format for ggplot
df_correct_long = df_correct %>%
  pivot_longer(cols = -Participant, names_to = "Drug", values_to = "Accuracy") %>%
  mutate(Classification = ifelse(Accuracy == 1, "Correct (Hit)", "Incorrect (Error)"))

# 6. Filter for the first 50 participants 
df_correct_subset = df_correct_long %>% filter(Participant <= 50)

# 7. Create the Evaluation Heatmap
ggplot(data = df_correct_subset, aes(x = Drug, y = factor(Participant), fill = Classification)) +
  geom_tile(color = "white", size = 0.1) + 
  scale_fill_manual(values = c("Correct (Hit)" = "#2ecc71", "Incorrect (Error)" = "#e74c3c")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    panel.grid = element_blank()
  ) +
  labs(
    tag = "Figure 8",
    title = "Model Prediction Accuracy Matrix",
    subtitle = "Ground Truth vs. MLP Binary Predictions (First 50 Test Participants)",
    x = "Substance",
    y = "Participant ID",
    fill = "Model Outcome"
  )
```

  How could our model be used in a clinical setting? We construct a synthetic profile of a new respondent to the survey and examine how the model predicts the probability of drug usage across our 15 target substances.

```{r }
# 1. Define a brand-new respondent's data (All 12 quantified features)
# Note: These values must be the quantified/scaled numbers matching the dataset #format!
new_respondent = data.frame(
  nscore    =  0.13606,  # Neuroticism
  escore    = -0.43999,  # Extraversion
  oscore    =  0.29338,  # Openness
  ascore    = -0.15487,  # Agreeableness
  cscore    = -0.52745,  # Conscientiousness
  impuslive =  0.19268,  # Impulsivity
  ss        =  0.40148   # Sensation Seeking
)

# 2. Convert to a 1-row matrix so Keras can read it
new_respondent_matrix <- as.matrix(new_respondent)

new_response_pred_probs = predict(mlp_model, new_respondent_matrix, verbose = 0)

# 3. Create an empty data frame to hold the predictions
single_predictions <- data.frame(
  Drug = target_drugs,
  Probability = round(as.numeric(new_response_pred_probs[1, ]), 4)
)

single_predictions_sorted <- single_predictions %>%
  arrange(desc(Probability))

# 6. View the ordered risk profile
print(single_predictions_sorted)
```
## RNN Model
```{r}
# Reshaping the data for RNN
X_train_rnn = array_reshape(as.matrix(X_train), c(nrow(X_train), 7, 1))
X_test_rnn = array_reshape(as.matrix(X_test), c(nrow(X_test), 7, 1))

# RNN model
inputs = layer_input(shape = c(7, 1)) 
outputs = inputs %>%
  layer_lstm(units = 64) %>%  
  layer_dense(units = 15, activation = 'sigmoid')

rnn_model = keras_model(inputs = inputs, outputs = outputs)

# Evaluating the effectiveness of this model using F1 score

library(yardstick)

# Converting props to binary
pred_probs_rnn = predict(rnn_model, X_test_rnn, verbose = 0)
pred_binary_rnn = ifelse(pred_probs_rnn > 0.5, 1, 0)

# Calculating for each drug individually
eval_results_rnn = data.frame()

for( i in 1:ncol(y_test)){
  drug_name = target_drugs[i]
  temp_df = data.frame(
    truth = factor(y_test[, i], levels = c(0, 1)),
    estimate = factor(pred_binary_rnn[, i], levels = c(0, 1))
  )
  # Getting F1 score
  f1 = f_meas(temp_df, truth, estimate)
  eval_results_rnn = rbind(eval_results_rnn, data.frame(Drug = drug_name, F1 = f1$.estimate))
}
print(eval_results_rnn)

```
Now, let's compare the F1 scores achieved from each of the models. 
```{r}
library(ggplot2)
library(dplyr)

plot_data = rbind(
  mutate(eval_results, Model = "MLP"),
  mutate(eval_results_rnn, Model = "RNN")
)

ggplot(plot_data, aes(x = Drug, y = F1, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "F1-Score Comparison: MLP vs RNN", y = "F1 Score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


## Analysis and Results

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}

  

```

### Modeling and Results

  There were several preprocessing steps to make the data ready for the ANN model. We began by verifying that the data set did not need standardization. We then converted the drug columns (the targets) into a binary "user" or "non user" designation based on whether the participant has used this substance within the last year. The study also used a fake drug called "Semeron" that was used to identify participants erroneously marking drugs not actually used. We omitted any instances in the dataset where a participant recorded use of "Semer" as accuracy of the remaining disclosed drugs used could not be verified. 

  We then convert the features into a numeric matrix. As the goal of our model is to predict drug use on personality and demographic responses and not the use of other drugs, we must remove the other drugs from the numeric feature matrix. 

  Before moving onto train/test splitting the data, a class imbalance between many drug classes was observed. Applying a randomized train/test split can compound this imbalance issue and impact the model so we used a composite profile score. By using this method, individuals that use more than one drug will be evenly distributed betweeen the train and test groups. 

For preprocessing the data for RNN, we must convert the matrices used to arrays

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
