---
title: "Predicting Drug Use Using Personailty Type and Pyschological Diagnoses- Summer 2026 "
subtitle: "A comparasion between RNN and DNN on non-sequential data"
author: "Elizabeth Rose and Victoria Kipker (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
engine: knitr
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

<style>
body {
  text-align: justify;
}
</style>

## Introduction

In an era plagued with a severe mental health crisis accelerated by both socioeconomic strife and the cognitive overload of the information age, the temptation to cope or escape can be compelling. As individuals attempt to navigate the current global landscape, many seek assistance from unregulated sources online, such as social media, frequently leading to self-diagnosis and self-medication. This access to “over-information” and a wide variety of ever-evolving psychoactive substances available have essentially left individuals overwhelmed by choice. Consequently, a critical behavioral question remains: what innate factors influence an individual’s substance of choice?

To investigate this phenomenon, research conducted by Fehrman et al. explored individual’s risk for drug consumption using a dataset collected via an online survey. [@fehrman2017five] Administered over a 12-month period, the survey resulted in usable data from 1,885 participants across a range of English-speaking nations including the United Kingdom, the United States, Canada, Australia, New Zealand, and Ireland. In addition to collecting information regarding the use of 18 legal and illegal drugs, the survey collected demographic metrics, such as education level and race, as well as their personality traits.


Personality traits were assessed using the Revised NEO Five-Factor Inventory (NEO-FFI-R). This method evaluates personalities based on scores across five core dimensions: neuroticism, extraversion, openness to experience, agreeableness, and conscientiousness [@costa1992revised]. This dataset established by Fehrman et al. serves as the foundation for this study. This study attempts to model the complex relationship between personality profiles and an individual’s “drug of choice”, while simultaneously predicting substance risks by implementing both traditional Deep Neural Networks (DNNs), such as the multilayer perceptron, as well as Recurrent Neural Networks (RNNs). 


<!-- -->

What better tool to predict the decisions from a human brain than by an algorithm designed to work like one: the neural network. In our case, we explored both traditional Deep Neural Networks (DNNs) like multilayer perceptrons as well as Recurrent Neural Networks (RNNs). 

The concept for the function of artificial neural networks (ANNs) has always been to replicate the processing methods of the brain: a complex, multi-parallel system of individual processing units called “neurons” [@MontesinosLopez2022] [@Larochelle2009Exploring]. While the inception of the artificial neuron dates back to the early 1940s, the underlying mechanism has remained fundamentally unchanged. Early iterations of ANNs were structurally simple, consisting of a single input layer, activation function, and an output layer. At the time, these models were operating with both structural and computational limitations, and as such were restricted to solving simpler, linearly separable problems [@ekundayo2025deep]. The field then experienced a prolonged period of stagnation until it ended abruptly in 1986 with the introduction of back propagation. This optimization method allows networks to systematically and repeatedly adjust their weights across multiple hidden layers to minimize the difference between the actual output and the desired output [@Rumelhart1986]. In the decades since the popularization of back propagation, advancements in computing power and data availability have accelerated the evolution of deep learning, to the point where ANNs have permeated virtually all corners of modern data analysis. 

The appropriate "version" of ANN to select is largely based on the input data at hand. Are your data points depended on one another or dependent on time itself? An RNN is best suited to this processing. Do you intend to decipher and interpret imagery? Best look into CNNs. But why? And furthermore, what version should we use for this drug dataset? 

Traditional feed forward DNNs process data in a one way flow. A human brain, in contrast, is a highly interconnected network with feedback loops and memory, much like the structure of an RNN. Our model is attempting to replicate the function of the brain however, the data is completely composed of continuous or categorical non-sequential observations. To rectify this incongruity and leave nary a stone unturned, we explored both a traditional DNN and an RNN.


## Methods

The multilayer perceptron (MLP) remains the standard feedforward architecture for processing structured, tabular data. Within an MLP, information can only flow in a single direction from the input layer through to the hidden processing layer or layers. Notably, neurons within the same layer are not interconnected. [@elmandouh2026comparing]. While more advanced models have been developed, MLPs have remained the benchmark when both simplicity and computational efficiency are the primary objectives. Furthermore, many optimization techniques have been developed to fine-tune these models, significantly accelerating convergence rates during training [@przybyla2024multi].
	
The selection of an appropriate ANN is dictated by the structure of the input data. For time series data, a recurrent neural network (RNN) provides increased analytical power by incorporating short-term memory. Unlike the unidirectional flow within an MLP, an RNN allows information to propagate through feedback loops. This enables the neurons to have connections to the previous layer, connections to neurons in the subsequent layer, and connections that transmit information between other neurons within the same layer. Essentially, this built-in memory enables the model to maintain context across a sequence to link current inputs to historical information [@elmandouh2026comparing].

For the analysis of time series data, language data, or data with a spatial structure, such as images or videos, a convolutional neural network (CNN) is typically the most effective. These models function by applying a series of convolutional filters that perform hierarchical feature extraction. Initially, low-level details such as edges or points are analyzed, while subsequent layers analyze higher-level shapes and contours. Since CNNs utilize the spatial correlation between adjacent data points, data generally requires less preprocessing than more traditional methods [@elmandouh2026comparing]. 

For the purpose of this paper, our analysis will focus exclusively on MLP and RNN models. By comparing the feed-forward processing of the MLP against the feedback-oriented structure of the RNN, we aim to determine if personality traits are good predictors of illicit drug use. 


-   Detail the structure and mathematics of the ANN/RNN as well as activation functions/sigmoid functions or other algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

```{r}
library(readr)

# Reading the file
library(ucimlrepo)
drug_data = fetch_ucirepo(id=373)
data = drug_data$data$original
#head(data)  

# Demographic reatures have already been scaled (Z-score standardization), however drug columns need to be converted
# Rows where individuals claimed "semer" use need to be removed since it was included as a fake drug


library(dplyr)

# Vector of all the drug column names
drug_cols = c(
  "alcohol", "amphet", "amyl", "benzos", "caff", "cannabis", "choc", 
  "coke", "crack", "ecstasy", "heroin", "ketamine", "legalh", "lsd", 
  "meth", "mushrooms", "nicotine", "semer", "vsa"
)

target_drugs = c(
  "amphet", "amyl", "benzos", "cannabis",
  "coke", "crack", "ecstasy", "heroin", "ketamine", "legalh", "lsd", 
  "meth", "mushrooms", "nicotine", "vsa"
)

# Now we filter out "Semeron" participants by keeping ones that say "never used"
data_cleaned = data %>%
  filter(semer == "CL0")


# Converting drug consumption columns into a binary target (user vs non-user)
# 0(non-user) and 1 (user)
## We will say anyone who has used the particular drug within the last year is a "user"

data_all_binary = data_cleaned %>%
  mutate(across(
    .cols = all_of(drug_cols),
    .fns = ~ ifelse(. %in% c("CL3", "CL4", "CL5", "CL6"), 1, 0)
  ))


# Checking to make sure they are binary
#head(data_all_binary[, c("alcohol", "amphet", "amyl")])


# Now we need to convert the features to a numeric matrix (X)
# Since we only want to train on personality scores, not other drug use, we need to remove other drug columns, too. 
X_data = data_all_binary %>%
  select(nscore, escore, oscore, ascore, cscore, impuslive, ss) %>%
  as.matrix()
#dim(X_data)


# Extracting the target variables (drug columns)
y_matrix = data_all_binary %>%
  select(all_of(target_drugs)) %>%
  as.matrix()
# Verifying the shape
#print(paste("Features Matrix (X) shape:", paste(dim(X_data), collapse = "x")))
#print(paste("Multi-Label Target Matrix (y) shape:", paste(dim(y_matrix), collapse = "x")))
y_matrix1 = data_all_binary %>%
  select(all_of(drug_cols)) %>%
  as.matrix()

# Plotting class imbalance 
library(ggplot2)
library(tidyr)
library(dplyr)

# Converting matrix to a dataframe
imbalance_plot_data = as.data.frame(y_matrix) %>%
  gather (key = "Drug", value = "Status") %>%
  mutate (Status = ifelse(Status == 1, "User", "Non-User"))

imbalance_plot_data1 = as.data.frame(y_matrix1) %>%
  gather (key = "Drug", value = "Status") %>%
  mutate (Status = ifelse(Status == 1, "User", "Non-User"))

# Creating a bar chart
ggplot(imbalance_plot_data1, aes (x = reorder(Drug, Status == "User", FUN = sum), fill = Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Non-User"= "gray", "User" = "blue"))+
  coord_flip() +
  labs(
    title = "Class Imbalance Across All 18 Target Substances",
    x = "Substance",
    y = "Percentage of Dataset",
    fill = "Classification",
  ) +
  theme_minimal()

ggplot(imbalance_plot_data, aes (x = reorder(Drug, Status == "User", FUN = sum), fill = Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Non-User"= "gray", "User" = "darkgreen"))+
  coord_flip() +
  labs(
    title = "Class Imbalance Across 15 Target Substances",
    x = "Substance",
    y = "Percentage of Dataset",
    fill = "Classification",
  ) +
  theme_minimal()

```
Based on the above bar charts, we noticed two facts about the data: there is a large class imbalance in reported drugs used (very few respondents reporting usage of drugs like crack) and this may cause the model to struggle in predicting use of those, and there are three substances that most if not all respondents reported use of: choc (chocolate), caff (caffeine), and alcohol. This beccomes important when we attempt to rectify class imbalances before splitting into train/test groups. 

A random split without taking into account this class imbalance can compound the imbalance impact. In order to mitigate this, a composite score is used. A sum of substances used is calculated for each of the instances in the dataset then taken into consideration for the tran/test split purposes (this results in respondents remporting multiple substances used are balanced between train and test). 

The purpose in leaving out the three universal baseline substances attempts to make the composite score more reflective of illicit or multi drug use. 
```{r}
# Since we have imbalance for the majority of the drugs, we need to be careful with out train/test split. 
# A random split will just compound the imbalance issue.
# We will use a composite profile score 
library(caret)
set.seed(10)

# By using this method, individuals that use more than one drug will be evenly distributed
composite_score = as.factor(rowSums(y_matrix))

# Conducting the 80/20 split
train_indices = createDataPartition(composite_score, p = 0.80, list = FALSE)

X_train = X_data[train_indices,]
X_test = X_data[-train_indices,]
y_train = y_matrix[train_indices,]
y_test = y_matrix[-train_indices,]

# Testing out the dimensions 
#print(paste("X_train:", paste(dim(X_train), collapse = " x ")))
#print(paste("y_train:", paste(dim(y_train), collapse = " x ")))
#print(paste("X_test:", paste(dim(X_test), collapse = " x ")))
#print(paste("y_test:", paste(dim(y_test), collapse = " x ")))

```

```{r}
# 1. Calculate the user percentages for the training set
train_dist <- as.data.frame(y_train) %>%
  gather(key = "Drug", value = "Status") %>%
  group_by(Drug) %>%
  summarize(User_Percentage = mean(Status == 1) * 100) %>%
  mutate(Split = "Train")

# 2. Calculate the user percentages for the test set
test_dist <- as.data.frame(y_test) %>%
  gather(key = "Drug", value = "Status") %>%
  group_by(Drug) %>%
  summarize(User_Percentage = mean(Status == 1) * 100) %>%
  mutate(Split = "Test")

# 3. Combine both distributions into a single dataframe
split_comparison <- rbind(train_dist, test_dist)

# 4. Create the side-by-side bar chart
ggplot(split_comparison, aes(x = reorder(Drug, User_Percentage), y = User_Percentage, fill = Split)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("Train" = "#2c3e50", "Test" = "#e74c3c")) +
  coord_flip() +
  labs(
    title = "Distribution of Drug Users Across Train and Test Splits",
    subtitle = "Verifying stratification balance (Omitting Alcohol, Caffeine, and Chocolate)",
    x = "Substance",
    y = "Percentage of Users (%)",
    fill = "Dataset Split"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 10)
  )
```


```{r}
library(reshape2)
library(igraph)
library(ggnetwork)
library(ggplot2)

# 1. Calculate correlation matrix between all 15 remaining drugs
drug_corr <- cor(y_matrix) 
diag(drug_corr) <- 0 

# 2. Filter out weak connections
drug_corr[abs(drug_corr) < 0.25] <- 0 

# 3. Build the network object
network_structure <- graph_from_adjacency_matrix(drug_corr, weighted = TRUE, mode = "undirected")

# 4. SPREAD OUT THE NETWORK: Generate spaced coordinates using a scaling trick
set.seed(123) # For reproducibility

# We pass an inverse weight array (1 - correlation) or a fixed small weight 
# to force the Fruchterman-Reingold layout to push nodes further apart
spaced_layout <- layout_with_fr(
  network_structure, 
  niter = 1500, 
  weights = rep(0.1, ecount(network_structure)) # Lower weight = weaker springs = more spread out!
)

# 5. Convert to ggnetwork using our custom spaced coordinates
network_data <- ggnetwork(network_structure, layout = spaced_layout)

# 6. Plot the spacious layout
ggplot(network_data, aes(x = x, y = y, xend = xend, yend = yend)) +
  
  # CHANGES HERE: Map link color to the correlation weight and make lines thicker
  geom_edges(aes(color = weight), size = 1.2, alpha = 0.8) +
  
  # Gradient palette for the links (e.g., yellow to red for low to high correlation)
  scale_color_viridis_c(option = "plasma", name = "Correlation\nStrength", direction = -1) +
  
  # Uniform node coloring (a clean slate grey or dark blue so the links pop)
  geom_nodes(color = "gray", size = 6) +
  
  # Labels positioned neatly over the nodes
  geom_nodetext(aes(label = name), fontface = "bold", color = "black", size = 4) +
  
  theme_blank() +
  labs(
    title = "The Web of Substance Co-occurrence",
    subtitle = "Link colors reflect correlation strength (Threshold > 0.25)"
  ) +
  theme(
    legend.position = "right"
  )


```
From this web we illustrate the strong ties between use of multiple substances. As the lines increase in darkness, the strenght of the usage correlation increases. 

Let's look at the "psychological signature" for some of these substances

```{r}
library(dplyr)
library(tidyr)
# 1. Install remotes if you don't already have it
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# 2. Download and install ggradar from GitHub
remotes::install_github("ricardo-bion/ggradar")
library(ggradar) 

min_max_scale <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

# 2. Scale the raw respondent traits globally FIRST
data_scaled_global <- data_all_binary %>%
  mutate(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), min_max_scale))

# 1. Prepare aggregate averages for cannabis
footprint_cannabis <- data_scaled_global %>%
  group_by(cannabis) %>%
  summarise(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), mean)) %>%
  rename(group = cannabis) %>%
  mutate(group = ifelse(group == 1, "Cannabis User", "Non-User"))

colnames(footprint_cannabis) <- c("group", "Neuroticism", "Extraversion", "Openness", "Agreeableness", "Conscientiousness", "Impulsivity", "Sensation_Seeking")

# 3. Generate the Radar Chart explicitly using the new meth object
ggradar(
  plot.data = footprint_cannabis, 
  grid.min = 0, grid.mid = 0.5, grid.max = 0.7, # Adjust grid boundaries to fit the new global means
  group.point.size = 3, group.line.width = 1, legend.position = "bottom"
) + 
  labs(title = "The Psychological Footprint of Cannabis Consumption")
```
Examining the resulting polygons shows us that a user of cannabis is more likely to exhibit traits of sensation seeking and openness more than that reported in non-users. 

Let's examine a "more illicit substance", Crack

```{r}
min_max_scale <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

# 2. Scale the raw respondent traits globally FIRST
data_scaled_global <- data_all_binary %>%
  mutate(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), min_max_scale))

# 1. Prepare aggregate averages for Meth
footprint_crack <- data_scaled_global %>%
  group_by(crack) %>%
  summarise(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), mean)) %>%
  rename(group = crack) %>%
  mutate(group = ifelse(group == 1, "Crack User", "Non-User"))

colnames(footprint_crack) <- c("group", "Neuroticism", "Extraversion", "Openness", "Agreeableness", "Conscientiousness", "Impulsivity", "Sensation_Seeking")

# 3. Generate the Radar Chart explicitly using the new meth object
ggradar(
  plot.data = footprint_crack, 
  grid.min = 0, grid.mid = 0.5, grid.max = 0.7, # Adjust grid boundaries to fit the new global means
  group.point.size = 3, group.line.width = 1, legend.position = "bottom"
) + 
  labs(title = "The Psychological Footprint of Crack Consumption")
```
Examining this polygon in contrast shows a larger disparity between non-users and users for the sensation seeking and impulsivity categories with a lower-than-non-user response for concientiousness and agreeableness. 


###Training a model to predict drug use on personality features
The below plots of accuracy and loss over epoch show that there is not a tremendous accuracy for our model. This is expected due to the number of different drug classes we are trying to generate probabilties for in one model. 
```{r}
######## READY FOR MLP USE WITH THIS TRAIN/TEST DATA########

library(keras3)
library(tensorflow)
# install.packages("keras3")


# Input layer
inputs = layer_input(shape = c(7))

outputs = inputs %>%
  layer_dense(units = 64, activation = 'relu') %>%
  layer_dropout(rate = 0.3) %>%
  layer_dense(units = 32, activation = 'relu') %>%
  layer_dropout(rate = 0.2) %>%
  layer_dense(units = 15, activation = 'sigmoid')

mlp_model = keras_model(input = inputs, output = outputs)

# Compiling the model
mlp_model %>% compile(
  optimizer = 'adam',
  loss = 'binary_crossentropy',
  metrics = c('accuracy')
)

# Training the model
history = mlp_model %>% fit(
  x = X_train,
  y = y_train,
  epochs = 20, 
  batch_size = 32,
  validation_split = 0.2,
  verbose = 0
)
  
plot(history)
```
We can instead examine the F1 scores of each drug individually to see how well the model is classifying users vs non-users. 
```{r}
# Evaluating the effectiveness of this model using F1 score
library(yardstick)
# Converting props to binary
pred_probs = predict(mlp_model, X_test, verbose = 0)
pred_binary = ifelse(pred_probs > 0.5, 1, 0)
results = data.frame(truth = as.matrix(y_test, estimate = pred_binary))

# Calculating for each drug individually
eval_results = data.frame()

for( i in 1:ncol(y_test)){
  drug_name = target_drugs[i]
  temp_df = data.frame(
    truth = factor(y_test[, i], levels = c(0, 1)),
    estimate = factor(pred_binary[, i], levels = c(0, 1))
  )
  # Getting F1 score
  f1 = f_meas(temp_df, truth, estimate)
  eval_results = rbind(eval_results, data.frame(Drug = drug_name, F1 = f1$.estimate))
}
print(eval_results)
```
Values close to 1 mean the model is predicting drug usage well. It is interesting to note that the model struggles the most with predicting nicotine users. 

Now that we have generated a model for predicting drug use, let's examine a heat map of participants and their predicted drugs used. 
```{r}

# Heat map for the first participant's likelihood profile
all_probs = predict(mlp_model, X_test, verbose = 0)

colnames(all_probs) = target_drugs

# Creating a data frame
df_probs = as.data.frame(all_probs)
df_probs$Participant = 1:nrow(df_probs)

df_long = df_probs %>%
  pivot_longer(cols = -Participant, names_to = "Drug", values_to = "Probability")

# Selecting the first 50 participants 
df_subset = df_long %>% filter(Participant <=50)

# Creating the heatmap
ggplot(data = df_subset,aes(x = Drug, y = factor(Participant), fill = Probability)) +
  geom_tile()+
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Individual Drug Usage Likelihoods",
       x = "Drug Category",
       y = "Participant ID")


```
And below we can see how the model performed, struggling the most with the nicotine category, as evidenced in the F1 scores. 
```{r}
# 1. Convert probabilities to binary predictions (using 0.5 threshold)
all_preds = ifelse(all_probs >= 0.5, 1, 0)
colnames(all_preds) = target_drugs

# 2. FIX: Ensure y_test matches all_preds in dimension and order
# Convert y_test to a clean matrix and align columns with your target drugs
y_test_matrix = as.matrix(y_test[, target_drugs])

# Double check rows: If y_test is still the full dataset, slice it to match the test size
if (nrow(y_test_matrix) != nrow(all_preds)) {
  # This grabs the trailing rows matching your test set size
  y_test_matrix = tail(y_test_matrix, nrow(all_preds)) 
}

# 3. Calculate correctness element-by-element safely
is_correct = (all_preds == y_test_matrix)

# 4. Convert to data frame and add Participant IDs
df_correct = as.data.frame(is_correct)
df_correct$Participant = 1:nrow(df_correct)

# 5. Reshape to long format for ggplot
df_correct_long = df_correct %>%
  pivot_longer(cols = -Participant, names_to = "Drug", values_to = "Accuracy") %>%
  mutate(Classification = ifelse(Accuracy == 1, "Correct (Hit)", "Incorrect (Error)"))

# 6. Filter for the first 50 participants 
df_correct_subset = df_correct_long %>% filter(Participant <= 50)

# 7. Create the Evaluation Heatmap
ggplot(data = df_correct_subset, aes(x = Drug, y = factor(Participant), fill = Classification)) +
  geom_tile(color = "white", size = 0.1) + 
  scale_fill_manual(values = c("Correct (Hit)" = "#2ecc71", "Incorrect (Error)" = "#e74c3c")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Model Prediction Accuracy Matrix",
    subtitle = "Ground Truth vs. MLP Binary Predictions (First 50 Test Participants)",
    x = "Substance",
    y = "Participant ID",
    fill = "Model Outcome"
  )
```

```{r }
# 1. Define a brand-new respondent's data (All 12 quantified features)
# Note: These values must be the quantified/scaled numbers matching the dataset #format!
new_respondent = data.frame(
  nscore    =  0.13606,  # Neuroticism
  escore    = -0.43999,  # Extraversion
  oscore    =  0.29338,  # Openness
  ascore    = -0.15487,  # Agreeableness
  cscore    = -0.52745,  # Conscientiousness
  impuslive =  0.19268,  # Impulsivity
  ss        =  0.40148   # Sensation Seeking
)

# 2. Convert to a 1-row matrix so Keras can read it
new_respondent_matrix <- as.matrix(new_respondent)

new_response_pred_probs = predict(mlp_model, new_respondent_matrix, verbose = 0)

# 3. Create an empty data frame to hold the predictions
single_predictions <- data.frame(
  Drug = target_drugs,
  Probability = round(as.numeric(new_response_pred_probs[1, ]), 4)
)

single_predictions_sorted <- single_predictions %>%
  arrange(desc(Probability))

# 6. View the ordered risk profile
print(single_predictions_sorted)
```


```{r}

```


### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

There were several preprocessing steps to make the data ready for the ANN model. We began by verifying that the data set did not need standardization. We then converted the drug columns (the targets) into a binary "user" or "non user" designation based on whether the participant has used this substance within the last year. The study also used a fake drug called "Semer" that was used to identify participants erroneously marking drugs not actually used. We omitted any instances in the dataset where a participant recorded use of "Semer" as accuracy of the remaining disclosed drugs used could not be verified. 

We then convert the features into a numeric matrix. As the goal of our model is to predict drug use on personality and demographic responses and not the use of other drugs, we must remove the other drugs from the numeric feature matrix. 

Before moving onto train/test splitting the data, a class imbalance between many drug classes was observed. Applying a randomized train/test split can compound this imbalance issue and impact the model so we used a composite profile score. By using this method, individuals that use more than one drug will be evenly distributed betweeen the train and test groups. 

For preprocessing the data for RNN, we must convert the matrices used to arrays

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
