---
title: "Predicting Drug Use Using Personailty Type and Pyschological Diagnoses- Summer 2026 "
subtitle: "A comparasion between RNN and DNN on non-sequential data"
author: "Elizabeth Rose and Victoria Kipker (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
engine: knitr
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

<style>
body {
  text-align: justify;
}
</style>

## Introduction

In an era plagued with a severe mental health crisis accelerated by both socioeconomic strife and the cognitive overload of the information age, the temptation to cope or escape can be compelling. As individuals attempt to navigate the current global landscape, many seek assistance from unregulated sources online, such as social media, frequently leading to self-diagnosis and self-medication. This access to “over-information” and a wide variety of ever-evolving psychoactive substances available have essentially left individuals overwhelmed by choice. Consequently, a critical behavioral question remains: what innate factors influence an individual’s substance of choice?

To investigate this phenomenon, research conducted by Fehrman et al. explored individual’s risk for drug consumption using a dataset collected via an online survey. [@fehrman2017five] Administered over a 12-month period, the survey resulted in usable data from 1,885 participants across a range of English-speaking nations including the United Kingdom, the United States, Canada, Australia, New Zealand, and Ireland. In addition to collecting information regarding the use of 18 legal and illegal drugs, the survey collected demographic metrics, such as education level and race, as well as their personality traits.


Personality traits were assessed using the Revised NEO Five-Factor Inventory (NEO-FFI-R). This method evaluates personalities based on scores across five core dimensions: neuroticism, extraversion, openness to experience, agreeableness, and conscientiousness [@costa1992revised]. This dataset established by Fehrman et al. serves as the foundation for this study. This study attempts to model the complex relationship between personality profiles and an individual’s “drug of choice”, while simultaneously predicting substance risks by implementing both traditional Deep Neural Networks (DNNs), such as the multilayer perceptron, as well as Recurrent Neural Networks (RNNs). 


<!-- -->

What better tool to predict the decisions from a human brain than by an algorithm designed to work like one: the neural network. In our case, we explored both traditional Deep Neural Networks (DNNs) like multilayer perceptrons as well as Recurrent Neural Networks (RNNs). 

The concept for the function of artificial neural networks (ANNs) has always been to replicate the processing methods of the brain: a complex, multi-parallel system of individual processing units called “neurons” [@MontesinosLopez2022]. While the inception of the artificial neuron dates back to the early 1940s, the underlying mechanism has remained fundamentally unchanged. Early iterations of ANNs were structurally simple, consisting of a single input layer, activation function, and an output layer. At the time, these models were operating with both structural and computational limitations, and as such were restricted to solving simpler, linearly separable problems [@ekundayo2025deep]. The field then experienced a prolonged period of stagnation until it ended abruptly in 1986 with the introduction of back propagation. This optimization method allows networks to systematically and repeatedly adjust their weights across multiple hidden layers to minimize the difference between the actual output and the desired output [@Rumelhart1986]. In the decades since the popularization of back propagation, advancements in computing power and data availability have accelerated the evolution of deep learning, to the point where ANNs have permeated virtually all corners of modern data analysis. 

The appropriate "version" of ANN to select is largely based on the input data at hand. Are your data points depended on one another or dependent on time itself? An RNN is best suited to this processing. Do you intend to decipher and interpret imagery? Best look into CNNs. But why? And furthermore, what version should we use for this drug dataset? 

Traditional feed forward DNNs process data in a one way flow. A human brain, in contrast, is a highly interconnected network with feedback loops and memory, much like the structure of an RNN. Our model is attempting to replicate the function of the brain however, the data is completely composed of continuous or categorical non-sequential observations. To rectify this incongruity and leave nary a stone unturned, we explored both a traditional DNN and an RNN.


## Methods

The basic structure of an artificial neural network is described as having an input layer followed by hidden layers, then a final output layer. What distinguishes an ANN from a DNN is the number of hidden layers.  A convolutional neural network processes patterned data (whether patterned as images or patterned by time or influence) through the inclusion of convolutional and pooling layers in order to identify features and compress the results. How an RNN is different: the inclusion of "memory" by incorporating a hidden state. In a neural network of this structure, each layer is considering the current input and the memory of its past inputs. Consider processing the information stored in a book. A traditional DNN takes in all words simultaneously, a CNN scans random clusters of pages, and an RNN reads each word sequentially keeping the context of the inputs. 

-   Detail the structure and mathematics of the ANN/RNN as well as activation functions/sigmoid functions or other algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

```{r}
library(readr)

# Defining the columns
column_names = c(
  "ID", "Age", "Gender", "Education", "Country", "Ethnicity", 
  "Nscore", "Escore", "Oscore", "Ascore", "Cscore", "Impulsive", "SS", 
  "Alcohol", "Amphet", "Amyl", "Benzos", "Caff", "Cannabis", "Choc", 
  "Coke", "Crack", "Ecstasy", "Heroin", "Ketamine", "Legalh", "LSD", 
  "Meth", "Mushroom", "Nicotine", "Semer", "VSA"
)

# Reading the file
library(ucimlrepo)
drug_data = fetch_ucirepo(id=373)
data = drug_data$data$original
#head(data)  

# Demographic reatures have already been scaled (Z-score standardization), however drug columns need to be converted
# Rows where individuals claimed "semer" use need to be removed since it was included as a fake drug


library(dplyr)

# Vector of all the drug column names
drug_cols = c(
  "alcohol", "amphet", "amyl", "benzos", "caff", "cannabis", "choc", 
  "coke", "crack", "ecstasy", "heroin", "ketamine", "legalh", "lsd", 
  "meth", "mushrooms", "nicotine", "semer", "vsa"
)

target_drugs = c(
  "amphet", "amyl", "benzos", "cannabis",
  "coke", "crack", "ecstasy", "heroin", "ketamine", "legalh", "lsd", 
  "meth", "mushrooms", "nicotine", "vsa"
)

# Now we filter out "Semeron" participants by keeping ones that say "never used"
data_cleaned = data %>%
  filter(semer == "CL0")


# Converting drug consumption columns into a binary target (user vs non-user)
# 0(non-user) and 1 (user)
## We will say anyone who has used the particular drug within the last year is a "user"

data_all_binary = data_cleaned %>%
  mutate(across(
    .cols = all_of(drug_cols),
    .fns = ~ ifelse(. %in% c("CL3", "CL4", "CL5", "CL6"), 1, 0)
  ))


# Checking to make sure they are binary
#head(data_all_binary[, c("alcohol", "amphet", "amyl")])


# Now we need to convert the features to a numeric matrix (X)
# Since we only want to train on demographic and personality scores, not other drug use, we need to remove other drug columns, too. 
X_data = data_all_binary %>%
  select(nscore, escore, oscore, ascore, cscore, impuslive, ss) %>%
  as.matrix()
#dim(X_data)


# Extracting the target variables (drug columns)
y_matrix = data_all_binary %>%
  select(all_of(target_drugs)) %>%
  as.matrix()
# Verifying the shape
#print(paste("Features Matrix (X) shape:", paste(dim(X_data), collapse = "x")))
#print(paste("Multi-Label Target Matrix (y) shape:", paste(dim(y_matrix), collapse = "x")))
y_matrix1 = data_all_binary %>%
  select(all_of(drug_cols)) %>%
  as.matrix()

# Plotting class imbalance 
library(ggplot2)
library(tidyr)
library(dplyr)

# Converting matrix to a dataframe
imbalance_plot_data = as.data.frame(y_matrix) %>%
  gather (key = "Drug", value = "Status") %>%
  mutate (Status = ifelse(Status == 1, "User", "Non-User"))

imbalance_plot_data1 = as.data.frame(y_matrix1) %>%
  gather (key = "Drug", value = "Status") %>%
  mutate (Status = ifelse(Status == 1, "User", "Non-User"))

# Creating a bar chart
ggplot(imbalance_plot_data1, aes (x = reorder(Drug, Status == "User", FUN = sum), fill = Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Non-User"= "gray", "User" = "blue"))+
  coord_flip() +
  labs(
    title = "Class Imbalance Across All 18 Target Substances",
    x = "Substance",
    y = "Percentage of Dataset",
    fill = "Classification",
  ) +
  theme_minimal()

ggplot(imbalance_plot_data, aes (x = reorder(Drug, Status == "User", FUN = sum), fill = Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Non-User"= "gray", "User" = "darkgreen"))+
  coord_flip() +
  labs(
    title = "Class Imbalance Across 15 Target Substances",
    x = "Substance",
    y = "Percentage of Dataset",
    fill = "Classification",
  ) +
  theme_minimal()

```

```{r}
# Since we have imbalance for the majority of the drugs, we need to be careful with out train/test split. 
# A random split will just compound the imbalance issue.
# We will use a composite profile score 
library(caret)
set.seed(10)

# By using this method, individuals that use more than one drug will be evenly distributed
composite_score = as.factor(rowSums(y_matrix))

# Conducting the 80/20 split
train_indices = createDataPartition(composite_score, p = 0.80, list = FALSE)

X_train = X_data[train_indices,]
X_test = X_data[-train_indices,]
y_train = y_matrix[train_indices,]
y_test = y_matrix[-train_indices,]

# Testing out the dimensions 
#print(paste("X_train:", paste(dim(X_train), collapse = " x ")))
#print(paste("y_train:", paste(dim(y_train), collapse = " x ")))
#print(paste("X_test:", paste(dim(X_test), collapse = " x ")))
#print(paste("y_test:", paste(dim(y_test), collapse = " x ")))


######## READY FOR MLP USE WITH THIS TRAIN/TEST DATA########

library(keras3)
library(tensorflow)
# install.packages("keras3")


# Input layer
inputs = layer_input(shape = c(7))

outputs = inputs %>%
  layer_dense(units = 64, activation = 'relu') %>%
  layer_dropout(rate = 0.3) %>%
  layer_dense(units = 32, activation = 'relu') %>%
  layer_dropout(rate = 0.2) %>%
  layer_dense(units = 15, activation = 'sigmoid')

mlp_model = keras_model(input = inputs, output = outputs)

# Compiling the model
mlp_model %>% compile(
  optimizer = 'adam',
  loss = 'binary_crossentropy',
  metrics = c('accuracy')
)

# Training the model
history = mlp_model %>% fit(
  x = X_train,
  y = y_train,
  epochs = 20, 
  batch_size = 32,
  validation_split = 0.2,
  verbose = 0
)
  
plot(history)

# Evaluating the effectiveness of this model using F1 score
library(yardstick)
# Converting props to binary
pred_probs = predict(mlp_model, X_test, verbose = 0)
pred_binary = ifelse(pred_probs > 0.5, 1, 0)
results = data.frame(truth = as.matrix(y_test, estimate = pred_binary))

# Calculating for each drug individually
eval_results = data.frame()

for( i in 1:ncol(y_test)){
  drug_name = target_drugs[i]
  temp_df = data.frame(
    truth = factor(y_test[, i], levels = c(0, 1)),
    estimate = factor(pred_binary[, i], levels = c(0, 1))
  )
  # Getting F1 score
  f1 = f_meas(temp_df, truth, estimate)
  eval_results = rbind(eval_results, data.frame(Drug = drug_name, F1 = f1$.estimate))
}
print(eval_results)
# Values close to 1 mean the model is predicting drug usage well. It is interesting to note that the model struggles  
# the most with predicting nicotine users. 

# Heat map for the first participant's likelihood profile
all_probs = predict(mlp_model, X_test, verbose = 0)

colnames(all_probs) = target_drugs

# Creating a data frame
df_probs = as.data.frame(all_probs)
df_probs$Participant = 1:nrow(df_probs)

df_long = df_probs %>%
  pivot_longer(cols = -Participant, names_to = "Drug", values_to = "Probability")

# Selecting the first 50 participants 
df_subset = df_long %>% filter(Participant <=50)

# Creating the heatmap
ggplot(data = df_subset,aes(x = Drug, y = factor(Participant), fill = Probability)) +
  geom_tile()+
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Individual Drug Usage Likelihoods",
       x = "Drug Category",
       y = "Participant ID")


```


```{r }
# 1. Define a brand-new respondent's data (All 12 quantified features)
# Note: These values must be the quantified/scaled numbers matching the dataset #format!
new_respondent = data.frame(
  nscore    =  0.13606,  # Neuroticism
  escore    = -0.43999,  # Extraversion
  oscore    =  0.29338,  # Openness
  ascore    = -0.15487,  # Agreeableness
  cscore    = -0.52745,  # Conscientiousness
  impuslive =  0.19268,  # Impulsivity
  ss        =  0.40148   # Sensation Seeking
)

# 2. Convert to a 1-row matrix so Keras can read it
new_respondent_matrix <- as.matrix(new_respondent)

new_response_pred_probs = predict(mlp_model, new_respondent_matrix, verbose = 0)

# 3. Create an empty data frame to hold the predictions
single_predictions <- data.frame(
  Drug = target_drugs,
  Probability = round(as.numeric(new_response_pred_probs[1, ]), 4)
)

single_predictions_sorted <- single_predictions %>%
  arrange(desc(Probability))

# 6. View the ordered risk profile
print(single_predictions_sorted)
```


```{r}

```


### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

There were several preprocessing steps to make the data ready for the ANN model. We began by verifying that the data set did not need standardization. We then converted the drug columns (the targets) into a binary "user" or "non user" designation based on whether the participant has used this substance within the last year. The study also used a fake drug called "Semer" that was used to identify participants erroneously marking drugs not actually used. We omitted any instances in the dataset where a participant recorded use of "Semer" as accuracy of the remaining disclosed drugs used could not be verified. 

We then convert the features into a numeric matrix. As the goal of our model is to predict drug use on personality and demographic responses and not the use of other drugs, we must remove the other drugs from the numeric feature matrix. 

Before moving onto train/test splitting the data, a class imbalance between many drug classes was observed. Applying a randomized train/test split can compound this imbalance issue and impact the model so we used a composite profile score. By using this method, individuals that use more than one drug will be evenly distributed betweeen the train and test groups. 

For preprocessing the data for RNN, we must convert the matrices used to arrays

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
