---
title: "Predicting Drug Use Using Personailty Type and Pyschological Diagnoses- Summer 2026 "
subtitle: "A comparasion between RNN and DNN on non-sequential data"
author: "Elizabeth Rose and Victoria Kipker (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)


## Introduction

In an era plagued with a severe mental health crisis accelerated by both socioeconomic strife and the cognitive overload of the information age, the temptation to cope or escape can be compelling. As individuals attempt to navigate the current global landscape, many seek assistance from unregulated sources online, such as social media, frequently leading to self-diagnosis and self-medication. This access to “over-information” and a wide variety of ever-evolving psychoactive substances available have essentially left individuals overwhelmed by choice. Consequently, a critical behavioral question remains: what innate factors influence an individual’s substance of choice?

To investigate this phenomenon, research conducted by Fehrman et al. explored individual’s risk for drug consumption using a dataset collected via an online survey. [@fehrman2017five] Administered over a 12-month period, the survey resulted in usable data from 1,885 participants across a range of English-speaking nations including the United Kingdom, the United States, Canada, Australia, New Zealand, and Ireland. In addition to collecting information regarding the use of 18 legal and illegal drugs, the survey collected demographic metrics, such as education level and race, as well as their personality traits.


Personality traits were assessed using the Revised NEO Five-Factor Inventory (NEO-FFI-R). This method evaluates personalities based on scores across five core dimensions: neuroticism, extraversion, openness to experience, agreeableness, and conscientiousness [@costa1992revised]. This dataset established by Fehrman et al. serves as the foundation for this study. This study attempts to model the complex relationship between personality profiles and an individual’s “drug of choice”, while simultaneously predicting substance risks by implementing both traditional Deep Neural Networks (DNNs), such as the multilayer perceptron, as well as Recurrent Neural Networks (RNNs). 


<!-- -->

What better tool to predict the decisions from a human brain than by an algorithm designed to work like one: the neural network. In our case, we explored both traditional Deep Neural Networks (DNNs) like multilayer perceptrons as well as Recurrent Neural Networks (RNNs). 

The concept for the function of artificial neural networks (ANNs) has always been to replicate the processing methods of the brain: a complex, multi-parallel system of individual processing units called “neurons” [@MontesinosLopez2022]. While the inception of the artificial neuron dates back to the early 1940s, the underlying mechanism has remained fundamentally unchanged. Early iterations of ANNs were structurally simple, consisting of a single input layer, activation function, and an output layer. At the time, these models were operating with both structural and computational limitations, and as such were restricted to solving simpler, linearly separable problems [@ekundayo2025deep]. The field then experienced a prolonged period of stagnation until it ended abruptly in 1986 with the introduction of back propagation. This optimization method allows networks to systematically and repeatedly adjust their weights across multiple hidden layers to minimize the difference between the actual output and the desired output [@Rumelhart1986]. In the decades since the popularization of back propagation, advancements in computing power and data availability have accelerated the evolution of deep learning, to the point where ANNs have permeated virtually all corners of modern data analysis. 

The appropriate "version" of ANN to select is largely based on the input data at hand. Are your data points depended on one another or dependent on time itself? An RNN is best suited to this processing. Do you intend to decipher and interpret imagery? Best look into CNNs. But why? And furthermore, what version should we use for this drug dataset? 

The basic structure of an artificial neural network is described as having an input layer followed by hidden layers, then a final output layer. What distinguishes an ANN from a DNN is the number of hidden layers. 

A convolutional neural network processes patterned data (whether patterned as images or patterned by time or influence) through the inclusion of convolutional and pooling layers in order to identify features and compress the results. 

How an RNN is different: the inclusion of "memory" by incorporating a hidden state. In a neural network of this structure, each layer is considering the current input and the memory of its past inputs. 

Consider processing the information stored in a book. A traditional DNN takes in all words simultaneously, a CNN scans random clusters of pages, and an RNN reads each word sequentially keeping the context of the inputs. 


Traditional feed forward DNNs process data in a one way flow. A human brain, in contrast, is a highly interconnected network with feedback loops and memory, much like the structure of an RNN. Our model is attempting to replicate the function of the brain however, the data is completely composed of continuous or categorical non-sequential observations. To rectify this incongruity and leave nary a stone unturned, we explored both a traditional DNN and an RNN.


## Methods


discuss preprocessing, any modifications or supplements to the data in case of class imbalance or whatever 


-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
