---
title: "Predicting Drug Use Using Personailty Type and Pyschological Diagnoses- Summer 2026 "
subtitle:  "A comparasion between RNN and DNN on non-sequential data"
author: "Elizabeth Rose and Victoria Kipker (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::

## Introduction  {.smaller}

Navigating the Modern Crisis:
- Current mental health crisis accelerated by socioeconomic challenges and cognitive overload
- Coping mechanisms or methods of escape are enticing
- Many turn to unregulated online sources such as social media for self-diagnosis and medication
- Available drugs are ever-evolving
- Individuals are essentially overwhelmed by choice

## Our Goal

- To determine if personality type is a good indicator for an individual's substance of choice

- If personality is a good indicator, can a model be used to identify "at risk" individuals in a way that does not consider use of different substances to be mutually exclusive?


## Methods

We will compare a multilayer perceptron (MLP) model to a Recurrent neural network (RNN) to see which model can better predict substance use from personality type.

## MLP

- Standard feed-forward architecture for structured, tabular data
- Information flows in one direction
- Neurons within same layer are not interconnected
- Not the most advanced, but remained the benchmark when simplicity and efficiency are needed

## RNN

- Typically used for time series data
- Incorporates short-term memory
- Allows information to propagate through feedback loops
- Neurons within same layer are interconnected
- Model can maintain context across a sequence to link current and historical information 

## Structure of Our Models

MLP
- Input layer, 2 hidden layers, and 1 output layer
  - First hidden layer: 64 neurons with Rectified Linear Unit (ReLU) activation
  - Second hidden layer: 32 neurons with ReLU activation
  - Output layer: 15 neurons with sigmoid activation (important for multi-label classification)
  
RNN
- Input layer, 1 hidden layer, and 1 output layer
  - Hidden layer: limited short-term memory (LSTM) layer of 64 neurons
  - Output layer: 15 neurons and a sigmoid activation function


## Data Collection

- Data collected via an online survey by Fehrman et al.
- Survey included questions on drug use (18 legal and illegal substances), frequency of drug use, personality traits, and demographic information
- Personality trait information collected using Revised NEO Five-Factor Inventory, Barratt Impulsiveness Scale, and Impulsive Sensation Seeking scale
- Initially yielded results from 2,051 respondents (final sample size was 1,885)
- Majority of respondents were from the UK (55.4%) and the US (29.5%)
- Included fictitious drug "Semeron" as a validity check

################Include table 1????#####################



-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

## Data Exploration and Visualization {.smaller}

A study was conducted to determine how...

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=F}
# Load Data
#kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
