---
title: "Predicting Drug Use Using Personailty Type and Pyschological Diagnoses- Summer 2026 "
subtitle:  "A comparasion between MLP and RNN on non-sequential data"
author: "Elizabeth Rose and Victoria Kipker (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs:
    css: styles.css  # If you already have a CSS file, or use the option below:
    theme: default
    # You can inject custom CSS directly into the header like this:
    include-in-header:
      - text: |
          <style>
            .title {
              font-size: 2.0em !important; 
            }
          </style>
course: Capstone Projects in Data Science
bibliography: references.bib
#always_allow_html: true
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction  

### Navigating the Modern Crisis:

- Current mental health crisis accelerated by socioeconomic challenges and cognitive overload
- Coping mechanisms or methods of escape are enticing
- Many turn to unregulated online sources, such as social media or AI, for self-diagnosis and self-medication
- Available drugs are ever-evolving
- Individuals are essentially overwhelmed by choice

## Our Goal  

- To determine if personality type is a good indicator for an individual's substance of choice

- If personality is a good indicator, can a model be used to identify "at-risk" individuals in a way that does not consider use of different substances to be mutually exclusive?


## Methods 

We will compare a traditional Artificial Neural Network, a multilayer perceptron (MLP) model, to a recurrent neural network (RNN) to see which model can better predict substance use from personality type.


## MLP  {.smaller}

- Standard feed-forward architecture for structured, tabular data
- Information flows in one direction 
- Neurons within same layer are not interconnected 
- Not the most advanced, but has remained the benchmark when simplicity and efficiency are needed


[@elmandouh2026comparing]
[@przybyla2024multi]


## RNN  {.smaller}

- Typically used for time series data
- Incorporates short-term memory
- Allows information to propagate through feedback loops
- Neurons within same layer are interconnected
- Model can maintain context across a sequence to link current and historical information


[@elmandouh2026comparing]



## Model Specifications  {.smaller}

### MLP {.smaller}

- Input layer, 2 hidden layers, and 1 output layer
  - First hidden layer: 64 neurons with Rectified Linear Unit (ReLU) activation
  - Second hidden layer: 32 neurons with ReLU activation
  - Output layer: 15 neurons with sigmoid activation (important for multi-label classification)
 
### RNN {.smaller}

- Input layer, 1 hidden layer, and 1 output layer
  - Hidden layer: limited short-term memory (LSTM) layer of 64 neurons
  - Output layer: 15 neurons and a sigmoid activation function
  
## Classification threshold  
  
- For both models, a dynamic threshold using the base use rate was used as our classification threshold.

- e.g. if a drug had a use rate of 10% in the dataset, only the top 10% of predicted users will be given the class "1"

- We also tried median and mean dynamic thresholds with less success


## Data Collection  {.smaller}

- Data collected via an online survey by Fehrman et al.
- Survey included questions on drug use (18 legal and illegal substances), frequency of drug use, personality traits, and demographic information
- Personality trait information collected using Revised NEO Five-Factor Inventory, Barratt Impulsiveness Scale, and Impulsive Sensation Seeking scale
- Initially yielded results from 2,051 respondents (final sample size was 1,885)
- Majority of respondents were from the UK (55.4%) and the US (29.5%)
- Included fictitious drug "Semeron" as a validity check.

[@fehrman2017five]


## Variables

Personality trait and drug use features collected by Fehrman et al. [@fehrman2017five] via online survey
```{r}

```

```{r}
#| echo: false
#| message: false
#| warning: false
library(tibble)
library(gt)
# Set R's seed
set.seed(10)

# Set environment variables and seeds for TensorFlow
library(reticulate)
py_run_string("
import os
os.environ['PYTHONHASHSEED'] = '10'
import random
random.seed(10)
import numpy as np
np.random.seed(10)
import tensorflow as tf
tf.random.set_seed(10)
")

library(keras)
#redundant just in case above doesnt set the correct variable
tensorflow::set_random_seed(10, disable_gpu = TRUE)
# 1. Define the tibble first so R knows what it is
variable_dictionary <- tibble(
  `Variable Name` = c(
    "nscore", "escore", "oscore", "ascore", "cscore", "impuslive", "ss",
    "alcohol", "amphet", "amyl", "benzos", "caff", "cannabis", "choc",
    "coke", "crack", "ecstasy", "heroin", "ketamine", "legalh", "lsd",
    "meth", "mushrooms", "nicotine", "semer", "vsa"
  ),
  `Description` = c(
    "Neuroticism score", "Extraversion score", "Openness to Experience score",
    "Agreeableness score", "Conscientiousness score", "Impulsiveness (BIS-11) score",
    "Sensation Seeking (ImpSS) score", "Alcohol usage frequency (1-7)",
    "Amphetamine usage frequency (1-7)", "Amyl nitrite usage frequency (1-7)",
    "Benzodiazepine usage frequency (1-7)", "Caffeine usage frequency (1-7)",
    "Cannabis usage frequency (1-7)", "Chocolate usage frequency (1-7)",
    "Cocaine usage frequency (1-7)", "Crack usage frequency (1-7)",
    "Ecstasy usage frequency (1-7)", "Heroin usage frequency (1-7)",
    "Ketamine usage frequency (1-7)", "Legal highs usage frequency (1-7)",
    "LSD usage frequency (1-7)", "Methadone usage frequency (1-7)",
    "Magic mushrooms usage frequency (1-7)", "Nicotine usage frequency (1-7)",
    "Semeron (ficticious drug) usage frequency (1-7)", "Volatile Substance Abuse usage frequency (1-7)"
  )
)

# 2. Now render the scrollable gt table
variable_dictionary %>%
  gt() %>%
  cols_label(
    `Variable Name` = md("**Variable Name**"),
    Description = md("**Description**")
  ) %>%
  tab_options(
    container.height = px(400), # Keeps the height constrained to slide limits
    table.width = pct(100)
  )
```


## Data Exploration and Visualization {.smaller}

Visual of class imbalance: Figure 1
```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
library(ggplot2)
library(tidyr)
library(dplyr)
library(readr)
library(ucimlrepo)
library(patchwork)
library(reshape2)
library(igraph)
library(ggnetwork)
library(ggradar) 
library(caret)
library(keras3)
library(tensorflow)
library(yardstick)

```

```{r, warning=FALSE, echo=F, message=FALSE}

drug_data = fetch_ucirepo(id=373)
data = drug_data$data$original
drug_cols = c(
  "alcohol", "amphet", "amyl", "benzos", "caff", "cannabis", "choc", 
  "coke", "crack", "ecstasy", "heroin", "ketamine", "legalh", "lsd", 
  "meth", "mushrooms", "nicotine", "semer", "vsa"
)

target_drugs = c(
  "amphet", "amyl", "benzos", "cannabis",
  "coke", "crack", "ecstasy", "heroin", "ketamine", "legalh", "lsd", 
  "meth", "mushrooms", "nicotine", "vsa"
)
data_cleaned = data %>%
  filter(semer == "CL0")
data_all_binary = data_cleaned %>%
  mutate(across(
    .cols = all_of(drug_cols),
    .fns = ~ ifelse(. %in% c("CL3", "CL4", "CL5", "CL6"), 1, 0)
  ))

X_data = data_all_binary %>%
  select(nscore, escore, oscore, ascore, cscore, impuslive, ss) %>%
  as.matrix()
y_matrix = data_all_binary %>%
  select(all_of(target_drugs)) %>%
  as.matrix()
y_matrix1 = data_all_binary %>%
  select(all_of(drug_cols)) %>%
  as.matrix()

imbalance_plot_data = as.data.frame(y_matrix) %>%
  gather (key = "Drug", value = "Status") %>%
  mutate (Status = ifelse(Status == 1, "User", "Non-User"))

imbalance_plot_data1 = as.data.frame(y_matrix1) %>%
  gather (key = "Drug", value = "Status") %>%
  mutate (Status = ifelse(Status == 1, "User", "Non-User"))

ggplot(imbalance_plot_data1, aes (x = reorder(Drug, Status == "User", FUN = sum), fill = Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Non-User"= "gray", "User" = "blue"))+
  coord_flip() +
  labs(
    tag = "Figure 1",
    title = "Class Imbalance Across All 18 Target Substances",
    x = "Substance",
    y = "Percentage of Dataset",
    fill = "Classification"
  ) +
  theme_minimal()
  

```

## Psychological footprints

Similarities noticed between polygons hints at correlation of substances
```{r, warning=FALSE, echo=F, message=FALSE}
min_max_scale = function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

# Scaling the raw respondent traits globally
data_scaled_global = data_all_binary %>%
  mutate(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), min_max_scale))

# Preparing aggregate averages for cannabis
footprint_cannabis = data_scaled_global %>%
  group_by(cannabis) %>%
  summarise(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), mean)) %>%
  rename(group = cannabis) %>%
  mutate(group = ifelse(group == 1, "Cannabis User", "Non-User"))

colnames(footprint_cannabis) <- c("group", "Neuroticism", "Extraversion", "Openness", "Agreeableness", "Conscientiousness", "Impulsivity", "Sensation_Seeking")

# Generating the Radar Chart explicitly using cannabis object
p1 = ggradar(
  plot.data = footprint_cannabis,
  grid.min = 0, grid.mid = 0.5, grid.max = 0.7,
  values.radar = c("0%", "50%", "70%"),
  group.point.size = 3, group.line.width = 1, legend.position = "bottom",
  
  axis.label.size = 3.5,     # Shrinks trait names (Neuroticism, Impulsivity, etc.)
  grid.label.size = 3,       # Shrinks the percentage labels (0%, 50%, 100%)
  legend.text.size = 9       # Shrinks the legend text at the bottom
) + 
  labs(
    title = "Cannabis Consumption", 
    tag = "Figure 4"
  ) +
  theme(
    plot.tag = element_text(size = 10, face = "bold"),           # Neat, small"Figure 4"
    plot.title = element_text(size = 12, face = "bold"),         # Small, crisp title
    legend.title = element_text(size = 9)                        # Small legend title
  )


# Scaling the raw respondent traits globally 
data_scaled_global = data_all_binary %>%
  mutate(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), min_max_scale))

# Preparing aggregate averages for crack
footprint_crack = data_scaled_global %>%
  group_by(crack) %>%
  summarise(across(c(nscore, escore, oscore, ascore, cscore, impuslive, ss), mean)) %>%
  rename(group = crack) %>%
  mutate(group = ifelse(group == 1, "Crack User", "Non-User"))

colnames(footprint_crack) = c("group", "Neuroticism", "Extraversion", "Openness", "Agreeableness", "Conscientiousness", "Impulsivity", "Sensation_Seeking")

# Generating the Radar Chart explicitly using the new crack object
p2= ggradar(
  plot.data = footprint_crack,
  grid.min = 0, grid.mid = 0.5, grid.max = 0.7,
  values.radar = c("0%", "50%", "70%"),
  group.point.size = 3, group.line.width = 1, legend.position = "bottom",
  
  axis.label.size = 3.5,     
  grid.label.size = 3,       
  legend.text.size = 9       
) + 
  labs(
    title = "Crack Consumption", 
    tag = "Figure 5"
  ) +
  theme(
    plot.tag = element_text(size = 10, face = "bold"),           # Neat, small "Figure 5"
    plot.title = element_text(size = 12, face = "bold"),         # Small, crisp title
    legend.title = element_text(size = 9)                        # Small legend title
  )
p1 + p2
```

## 
```{r, warning=FALSE, echo=F, message=FALSE}
drug_corr = cor(y_matrix) 
diag(drug_corr) = 0 

# Filtering out weak connections
drug_corr[abs(drug_corr) < 0.25] = 0 

# Building the network object
network_structure = graph_from_adjacency_matrix(drug_corr, weighted = TRUE, mode = "undirected")

# Generating spaced coordinates using a scaling trick
set.seed(123) 

# We pass an inverse weight array (1 - correlation) or a fixed small weight 
# to force the Fruchterman-Reingold layout to push nodes further apart
spaced_layout = layout_with_fr(
  network_structure, 
  niter = 1500, 
  weights = rep(0.1, ecount(network_structure)) # Lower weight = weaker springs = more spread out!
)

# Converting to ggnetwork using our custom spaced coordinates
network_data = ggnetwork(network_structure, layout = spaced_layout)

# Plotting the spacious layout
ggplot(network_data, aes(x = x, y = y, xend = xend, yend = yend)) +
  
  # Mapping link color to the correlation weight and make lines thicker
  geom_edges(aes(color = weight), size = 1.2, alpha = 0.8) +
  
  # Gradient palette for the links (e.g., yellow to red for low to high correlation)
  scale_color_viridis_c(option = "plasma", name = "Correlation\nStrength", direction = -1) +
  
  # Uniform node coloring (a clean slate grey or dark blue so the links pop)
  geom_nodes(color = "gray", size = 6) +
  
  # Labels positioned neatly over the nodes
  geom_nodetext(aes(label = name), fontface = "bold", color = "black", size = 4) +
  
  theme_blank() +
  labs(
    tag = "Figure 6",
    title = "The Web of Substance Co-occurrence",
    subtitle = "Link colors reflect correlation strength (Threshold > 0.25)"
  ) +
  theme(
    legend.position = "right"
  )
```
## Preprocessing {.smaller}
#### Performed by original researchers:

- Personality scores were quantified to continuous variables that approximate standard normal distribution 
- Removal of incomplete/unusable respondent data
- Removal of data from individuals that claimed "semeron" use

#### Performed prior to our study:

- Binarization of "users" vs. "non-users" for each substance
  - Use within the past year was considered a "user"
- Removal of 3 most used substances (chocolate, caffeine, alcohol)
- For RNN model: converted the 2D matrices (for MLP) into 3D arrays
  - Created a sequential feature extractor 
 

## The MLP

```{r, warning=FALSE, echo=F, message=FALSE}
set.seed(10)

# By using this method, individuals that use more than one drug will be evenly distributed
composite_score = as.factor(rowSums(y_matrix))

# Conducting the 80/20 split
train_indices = createDataPartition(composite_score, p = 0.80, list = FALSE)

X_train = X_data[train_indices,]
X_test = X_data[-train_indices,]
y_train = y_matrix[train_indices,]
y_test = y_matrix[-train_indices,]

inputs = layer_input(shape = c(7))

outputs = inputs %>%
  layer_dense(units = 64, activation = 'relu') %>%
  layer_dropout(rate = 0.3) %>%
  layer_dense(units = 32, activation = 'relu') %>%
  layer_dropout(rate = 0.2) %>%
  layer_dense(units = 15, activation = 'sigmoid')

mlp_model = keras_model(input = inputs, output = outputs)

# Compiling the model
mlp_model %>% compile(
  optimizer = 'adam',
  loss = 'binary_crossentropy',
  metrics = c('binary_accuracy')
)
history = mlp_model %>% fit(
  x = X_train,
  y = y_train,
  epochs = 20, 
  batch_size = 32,
  validation_split = 0.2,
  verbose = 0
)
plot(history)
```
## Hit or miss? 
```{r, warning=FALSE, echo=F, message=FALSE}
# Generate predicted probabilities

all_probs = predict(mlp_model, X_test, verbose = 0)
y_test_matrix = as.matrix(y_test[, target_drugs])

all_preds = all_probs # copy structure

for(i in 1:ncol(all_probs)) {
  drug_name = target_drugs[i]
  # 1. Find the actual proportion of users for this drug in the test set
  actual_use_rate = mean(y_test_matrix[, drug_name]) 
  
  # 2. Find the corresponding percentile threshold (e.g., if use rate is 10%, threshold is the 90th percentile)
  class_threshold = quantile(all_probs[, i], 1 - actual_use_rate)
  
  # 3. Classify based on this realistic, drug-specific threshold
  all_preds[, i] = ifelse(all_probs[, i] >= class_threshold, 1, 0)
}

# Ensure colnames are preserved from the target drugs
colnames(all_preds) = target_drugs

# Ensure y_test matches all_preds in dimension and order
# Convert y_test to a clean matrix and align columns with your target drugs


# Double checking rows
if (nrow(y_test_matrix) != nrow(all_preds)) {
  y_test_matrix = tail(y_test_matrix, nrow(all_preds)) 
}

# Calculating correctness element-by-element 
is_correct = (all_preds == y_test_matrix)

# Converting to dataframe and adding Participant IDs
df_correct = as.data.frame(is_correct)
df_correct$Participant = 1:nrow(df_correct)

# Reshaping to long format for ggplot
df_correct_long = df_correct %>%
  pivot_longer(cols = -Participant, names_to = "Drug", values_to = "Accuracy") %>%
  mutate(Classification = ifelse(Accuracy == 1, "Correct (Hit)", "Incorrect (Error)"))

# Filtering for the first 50 participants 
df_correct_subset = df_correct_long %>% filter(Participant <= 50)

# Creating the Evaluation Heatmap
ggplot(data = df_correct_subset, aes(x = Drug, y = factor(Participant), fill = Classification)) +
  geom_tile(color = "white", size = 0.1) + 
  scale_fill_manual(values = c("Correct (Hit)" = "#2ecc71", "Incorrect (Error)" = "#e74c3c")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    panel.grid = element_blank()
  ) +
  labs(
    tag = "Figure 8",
    title = "Model Prediction Accuracy Matrix",
    subtitle = "Ground Truth vs. MLP Binary Predictions (Column-Wise Median Thresholds)",
    x = "Substance",
    y = "Participant ID",
    fill = "Model Outcome"
  )
```


## The RNN
```{r, warning=FALSE, echo=F, message=FALSE}
X_train_rnn = array_reshape(as.matrix(X_train), c(nrow(X_train), 7, 1))
X_test_rnn = array_reshape(as.matrix(X_test), c(nrow(X_test), 7, 1))

# RNN model
inputs = layer_input(shape = c(7, 1)) 
outputs = inputs %>%
  layer_lstm(units = 64) %>%  
  layer_dense(units = 15, activation = 'sigmoid')

rnn_model = keras_model(inputs = inputs, outputs = outputs)

rnn_model %>% compile(
  optimizer = 'adam',
  loss = 'binary_crossentropy',
  metrics = c('binary_accuracy') # Tracks binary accuracy over epochs
)

# Fitting the model 
rnn_history = rnn_model %>% fit(
  x = X_train_rnn,
  y = y_train,
  epochs = 20,
  batch_size = 32,
  validation_split = 0.2,
  verbose = 0
)

#
plot(rnn_history)
```

## MLP vs RNN

```{r, warning=FALSE, echo=F, message=FALSE}
pred_probs = predict(mlp_model, X_test, verbose = 0)
pred_binary = ifelse(pred_probs > 0.5, 1, 0)
results = data.frame(truth = as.matrix(y_test, estimate = pred_binary))

# Calculating for each drug individually with dynamic thresholds
eval_results_mlp = data.frame()

for(i in 1:ncol(y_test)){
  drug_name = target_drugs[i]
  
  # 1. Calculate the actual base rate (proportion of 1s in the ground truth)
  actual_use_rate = mean(y_test[, i] == 1)
  
  # 2. Set the threshold so that ONLY the top percentage (matching the base rate) are predicted as 1s
  # If 10% of people use a drug, the threshold is the 90th percentile (1 - 0.10)
  current_threshold = quantile(pred_probs[, i], 1 - actual_use_rate) 
  
  # Converting to binary using this column's class-balanced threshold
  binary_preds = ifelse(pred_probs[, i] >= current_threshold, 1, 0)
  
  truth_vec    = factor(y_test[, i], levels = c(0, 1))
  estimate_vec = factor(binary_preds, levels = c(0, 1))
  
  # Calculating metrics 
  f1   = f_meas_vec(truth_vec, estimate_vec, event_level = "second")
  prec = precision_vec(truth_vec, estimate_vec, event_level = "second")
  rec  = recall_vec(truth_vec, estimate_vec, event_level = "second")
  
  eval_results_mlp = rbind(eval_results_mlp, data.frame(
    Drug = drug_name, 
    Threshold = as.numeric(current_threshold), # tracked to see what was used
    F1 = ifelse(is.na(f1), 0, f1),
    Precision = ifelse(is.na(prec), 0, prec),
    Recall = ifelse(is.na(rec), 0, rec)
  ))
}

pred_probs_rnn = predict(rnn_model, X_test_rnn, verbose = 0)
# Fixed typo: changed pred_probs to pred_probs_rnn
pred_binary_rnn = ifelse(pred_probs_rnn > 0.5, 1, 0)
results = data.frame(truth = as.matrix(y_test, estimate = pred_binary_rnn))

# Calculating for each drug individually with dynamic thresholds
eval_results_rnn = data.frame()

for(i in 1:ncol(y_test)){
  drug_name = target_drugs[i]
  
  # 1. Calculate the actual base rate (proportion of 1s in the ground truth)
  actual_use_rate = mean(y_test[, i] == 1)
  
  # 2. Set the threshold so that ONLY the top percentage (matching the base rate) are predicted as 1s
  current_threshold = quantile(pred_probs_rnn[, i], 1 - actual_use_rate) 
  
  # Converting to binary using this column's class-balanced threshold
  binary_preds = ifelse(pred_probs_rnn[, i] >= current_threshold, 1, 0)
  
  truth_vec    = factor(y_test[, i], levels = c(0, 1))
  estimate_vec = factor(binary_preds, levels = c(0, 1))
  
  # Calculating metrics 
  f1   = f_meas_vec(truth_vec, estimate_vec, event_level = "second")
  prec = precision_vec(truth_vec, estimate_vec, event_level = "second")
  rec  = recall_vec(truth_vec, estimate_vec, event_level = "second")
  
  eval_results_rnn = rbind(eval_results_rnn, data.frame(
    Drug = drug_name, 
    Threshold = as.numeric(current_threshold), # tracked to see what was used
    F1 = ifelse(is.na(f1), 0, f1),
    Precision = ifelse(is.na(prec), 0, prec),
    Recall = ifelse(is.na(rec), 0, rec)
  ))
}

plot_data = rbind(
  mutate(eval_results_mlp, Model = "MLP"),
  mutate(eval_results_rnn, Model = "RNN")
)

ggplot(plot_data, aes(x = Drug, y = F1, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(tag= "Figure 9", title = "F1-Score Comparison: MLP vs RNN", y = "F1 Score") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Conclusion

- Personality traits can be good indicators for substance usage risk when using ANN model frameworks
- Achieved overall binary accuracy rate of greater than 82% for both models
- Using adaptive median threshold for classification caused precision to suffer, inflating number of false positives, improved with base use threshold
- MLP is overall more efficient/accurate for multi-label classification when compared to RNN
- RNN able to model complex relationships between personality data (cocaine)

## References
